\documentclass[11pt,a4paper,openany]{book}

\makeatletter\def\input@path{{../_shared/}}\makeatother
\usepackage{course}

\begin{document}
\raggedbottom

\coursetitle{Algorithms I}{BA4}{2025--2026}

\tableofcontents
\bigskip

These notes cover the core algorithmic techniques of the BA4 Algorithms I course:
asymptotic analysis, sorting, divide-and-conquer, heaps, binary search trees, and
graph algorithms (max flow, matchings).

% =========================================================
\chapter{Algorithms Fundamentals}
% =========================================================

\section{Definition of an Algorithm}
\begin{defnbox}
An \textbf{algorithm} is a well-defined computational procedure that takes some value (or
set of values) as \emph{input} and produces some value (or set of values) as \emph{output}
in a finite number of steps.
\end{defnbox}

We analyse algorithms along two axes:
\begin{itemize}
  \item \textbf{Correctness} — does it always produce the right answer?
  \item \textbf{Efficiency} — how many resources (time, space) does it consume?
\end{itemize}

% ---------------------------------------------------------
\section{Asymptotic Complexity}
% ---------------------------------------------------------

\begin{defnbox}[title=Asymptotic notations]
Let $f, g : \mathbb{N} \to \mathbb{R}^+$.
\begin{itemize}
  \item $f \in O(g)$ \textbf{(upper bound)}: $\exists\, c, n_0 > 0$ s.t.\ $f(n) \leq c\,g(n)$ for all $n \geq n_0$
  \item $f \in \Omega(g)$ \textbf{(lower bound)}: $\exists\, c, n_0 > 0$ s.t.\ $f(n) \geq c\,g(n)$ for all $n \geq n_0$
  \item $f \in \Theta(g)$ \textbf{(tight bound)}: $f \in O(g)$ and $f \in \Omega(g)$
  \item $f \in o(g)$ \textbf{(strict upper)}: $\lim_{n\to\infty} f(n)/g(n) = 0$
  \item $f \in \omega(g)$ \textbf{(strict lower)}: $\lim_{n\to\infty} f(n)/g(n) = \infty$
\end{itemize}
\end{defnbox}

\begin{propbox}[title=Common growth rates (slowest to fastest)]
\[
  O(1) \subset O(\log n) \subset O(n) \subset O(n\log n) \subset O(n^2)
  \subset O(n^3) \subset O(2^n) \subset O(n!)
\]
\end{propbox}

\begin{rembox}
For the \textbf{Master Theorem}: given $T(n) = a\,T(n/b) + f(n)$ with $a \geq 1$, $b > 1$:
\begin{itemize}
  \item if $f(n) = O(n^{\log_b a - \eps})$ then $T(n) = \Theta(n^{\log_b a})$
  \item if $f(n) = \Theta(n^{\log_b a})$ then $T(n) = \Theta(n^{\log_b a} \log n)$
  \item if $f(n) = \Omega(n^{\log_b a + \eps})$ then $T(n) = \Theta(f(n))$
\end{itemize}
\end{rembox}

% ---------------------------------------------------------
\section{Sorting Algorithms}
% ---------------------------------------------------------

\subsection{Insertion Sort}

\begin{propbox}[title=Complexity]
Time: $\Theta(n^2)$ worst/average, $\Theta(n)$ best (already sorted). \quad Space: $O(1)$ in-place.
\end{propbox}

\begin{lstlisting}[language=Python]
def insertion_sort(A):
    for j in range(1, len(A)):
        key = A[j]
        i = j - 1
        while i >= 0 and A[i] > key:
            A[i + 1] = A[i]
            i -= 1
        A[i + 1] = key
\end{lstlisting}

\textbf{Loop invariant:} at the start of each iteration, \texttt{A[0..j-1]} is sorted.

\subsection{Merge Sort}

\begin{propbox}[title=Complexity]
Time: $\Theta(n \log n)$ in all cases. \quad Space: $O(n)$ auxiliary.
\end{propbox}

\begin{lstlisting}[language=Python]
def merge_sort(A, p, r):
    if p < r:
        q = (p + r) // 2
        merge_sort(A, p, q)
        merge_sort(A, q + 1, r)
        merge(A, p, q, r)   # merge two sorted halves
\end{lstlisting}

\textbf{Recurrence:} $T(n) = 2\,T(n/2) + \Theta(n)$ $\Rightarrow$ $T(n) = \Theta(n \log n)$ by Master Theorem (case 2).

% ---------------------------------------------------------
\section{Divide and Conquer}
% ---------------------------------------------------------

\begin{defnbox}[title=Paradigm]
\textbf{Divide} the problem into subproblems, \textbf{conquer} each recursively, then
\textbf{combine} the results.
\end{defnbox}

\subsection{Maximum Subarray}

Find indices $i \leq j$ maximising $\sum_{k=i}^{j} A[k]$.

\begin{propbox}[title=Complexity]
$T(n) = 2\,T(n/2) + \Theta(n)$ $\Rightarrow$ $\Theta(n \log n)$.
(Kadane's linear algorithm also exists: $\Theta(n)$.)
\end{propbox}

\begin{lstlisting}
FIND-MAXIMUM-SUBARRAY(A, low, high):
    if high == low: return (low, high, A[low])
    mid = (low + high) / 2
    left  = FIND-MAXIMUM-SUBARRAY(A, low, mid)
    right = FIND-MAXIMUM-SUBARRAY(A, mid+1, high)
    cross = FIND-MAX-CROSSING(A, low, mid, high)
    return the one with maximum sum
\end{lstlisting}

\subsection{Matrix Multiplication}

\subsubsection{Naive algorithm}
Standard $O(n^3)$ triple loop, or the recursive divide-and-conquer that splits each
$n \times n$ matrix into four $n/2 \times n/2$ blocks: $T(n) = 8\,T(n/2) + \Theta(n^2)$,
giving the same $\Theta(n^3)$ by Master Theorem (case 1).

\subsubsection{Strassen's algorithm}
\begin{propbox}[title=Complexity]
$T(n) = 7\,T(n/2) + \Theta(n^2)$ $\Rightarrow$ $\Theta(n^{\log_2 7}) \approx \Theta(n^{2.807})$.
\end{propbox}

Key idea: compute only 7 products of submatrices (instead of 8) using the following
auxiliary matrices $M_1, \ldots, M_7$, each a linear combination of blocks of $A$ and $B$:
\[
  M_1 = (A_{11}+A_{22})(B_{11}+B_{22}), \quad
  M_2 = (A_{21}+A_{22})B_{11}, \quad \ldots
\]
Then reconstruct $C_{11}, C_{12}, C_{21}, C_{22}$ from $M_1$--$M_7$.

\subsection{Karatsuba Multiplication}

Multiply two $n$-digit integers faster than the naive $O(n^2)$ schoolbook method.

Split: $x = x_H b^{n/2} + x_L$, $y = y_H b^{n/2} + y_L$.

Compute only 3 multiplications:
\[
  p_1 = x_H \cdot y_H, \quad
  p_2 = x_L \cdot y_L, \quad
  p_3 = (x_H + x_L)(y_H + y_L)
\]
then $x \cdot y = p_1 b^n + (p_3 - p_1 - p_2)\,b^{n/2} + p_2$.

\begin{propbox}[title=Complexity]
$T(n) = 3\,T(n/2) + \Theta(n)$ $\Rightarrow$ $\Theta(n^{\log_2 3}) \approx \Theta(n^{1.585})$.
\end{propbox}

% ---------------------------------------------------------
\section{Heaps}
% ---------------------------------------------------------

\begin{defnbox}[title=Max-heap property]
A \textbf{max-heap} is a nearly complete binary tree where every node satisfies
$A[\text{parent}(i)] \geq A[i]$.
Stored as an array: $\text{LEFT}(i)=2i$, $\text{RIGHT}(i)=2i+1$, $\text{PARENT}(i)=\lfloor i/2\rfloor$.
\end{defnbox}

\subsection{Heapify}
\begin{propbox}[title=Complexity] $O(\log n)$ — tree height. \end{propbox}

\begin{lstlisting}
MAX-HEAPIFY(A, i, n):
    l = 2i;  r = 2i + 1;  largest = i
    if l <= n and A[l] > A[largest]: largest = l
    if r <= n and A[r] > A[largest]: largest = r
    if largest != i:
        swap A[i] and A[largest]
        MAX-HEAPIFY(A, largest, n)
\end{lstlisting}

\subsection{Build-Heap}
Call \texttt{MAX-HEAPIFY} on all internal nodes bottom-up.
Though each call is $O(\log n)$, a tight analysis gives $O(n)$ total.

\subsection{Heapsort}
\begin{propbox}[title=Complexity]
Time: $\Theta(n \log n)$ in all cases. \quad Space: $O(1)$ in-place.
\end{propbox}

\begin{lstlisting}
HEAPSORT(A, n):
    BUILD-MAX-HEAP(A, n)
    for i = n downto 2:
        swap A[1] and A[i]
        MAX-HEAPIFY(A, 1, i - 1)
\end{lstlisting}

\subsection{Priority Queue}
\begin{propbox}
\begin{tabular}{ll}
\texttt{Maximum}     & $O(1)$ \\
\texttt{Extract-Max} & $O(\log n)$ \\
\texttt{Insert}      & $O(\log n)$ \\
\texttt{Increase-Key}& $O(\log n)$ \\
\end{tabular}
\end{propbox}

% ---------------------------------------------------------
\section{Binary Search Trees}
% ---------------------------------------------------------

\begin{defnbox}[title=BST property]
For every node $x$: all keys in the left subtree $\leq x.\mathit{key} \leq$ all keys in
the right subtree.
\end{defnbox}

All basic operations run in $O(h)$ where $h$ is the height.
For a \emph{random} BST, $\mathbb{E}[h] = O(\log n)$; in the worst case $h = n-1$.

\begin{lstlisting}
TREE-SEARCH(x, k):
    if x == NIL or k == x.key: return x
    if k < x.key: return TREE-SEARCH(x.left, k)
    else:          return TREE-SEARCH(x.right, k)
\end{lstlisting}

\begin{propbox}[title=Operations — all $O(h)$]
Search, Minimum, Maximum, Successor, Predecessor, Insert, Delete.
\end{propbox}

\begin{warnbox}
A BST is only efficient when \emph{balanced}. For guaranteed $O(\log n)$, use a
self-balancing variant such as a \textbf{Red-Black tree} (height $\leq 2\log_2(n+1)$).
\end{warnbox}

% ---------------------------------------------------------
\section{Data Structures Comparison}
% ---------------------------------------------------------

\begin{center}
\begin{tabular}{lllll}
\toprule
Structure    & Search    & Insert      & Delete      & Notes \\
\midrule
Unsorted array & $O(n)$  & $O(1)$      & $O(n)$      & simple \\
Sorted array   & $O(\log n)$ & $O(n)$  & $O(n)$      & binary search \\
Linked list    & $O(n)$  & $O(1)$      & $O(1)^*$    & $^*$with pointer \\
BST            & $O(h)$  & $O(h)$      & $O(h)$      & $h = O(n)$ worst \\
Red-Black tree & $O(\log n)$ & $O(\log n)$ & $O(\log n)$ & balanced \\
Heap           & $O(n)$  & $O(\log n)$ & $O(\log n)$ & max/min in $O(1)$ \\
Hash table     & $O(1)^*$& $O(1)^*$    & $O(1)^*$    & $^*$expected \\
\bottomrule
\end{tabular}
\end{center}

% =========================================================
\chapter{Graph Algorithms — Max Flow}
% =========================================================

\section{Flow Networks}

\begin{defnbox}[title=Flow network]
A directed graph $G = (V, E)$ with:
\begin{itemize}
  \item a \textbf{source} $s$ and a \textbf{sink} $t$,
  \item a \textbf{capacity function} $c : V \times V \to \mathbb{R}_{\geq 0}$ (0 for non-edges),
  \item no antiparallel edges (replace $(u,v)$ and $(v,u)$ by splitting with an intermediate node).
\end{itemize}
A \textbf{flow} $f$ must satisfy capacity ($0 \leq f(u,v) \leq c(u,v)$) and conservation
($\sum_v f(v,u) = \sum_v f(u,v)$ for all $u \neq s, t$).
\end{defnbox}

The \textbf{value} of a flow is $|f| = \sum_v f(s,v) - \sum_v f(v,s)$.

\section{Ford-Fulkerson Method}

\begin{defnbox}[title=Residual network]
Given flow $f$, the residual network $G_f$ has residual capacity
$c_f(u,v) = c(u,v) - f(u,v)$ for forward edges and $c_f(v,u) = f(u,v)$ for back edges.
An \textbf{augmenting path} is any $s$-$t$ path in $G_f$.
\end{defnbox}

\begin{lstlisting}
FORD-FULKERSON(G, s, t):
    initialise f = 0 on all edges
    while exists augmenting path p in G_f:
        cf(p) = min capacity along p
        augment f along p by cf(p)
    return f
\end{lstlisting}

\begin{thmbox}[title=Max-flow min-cut theorem]
The following are equivalent for a flow $f$:
\begin{enumerate}
  \item $f$ is a maximum flow.
  \item There is no augmenting path in $G_f$.
  \item $|f|$ equals the capacity of some minimum cut $(S, T)$.
\end{enumerate}
\end{thmbox}

\section{Applications}

\subsection{Bipartite matching via max flow}

\begin{thmbox}[title=Algorithm]
\textbf{Goal:} Maximum matching in a bipartite graph $G = (L \cup R, E)$.

\textbf{Reduction:}
\begin{itemize}
  \item Add source $s$ with edge $s \to u$ (capacity 1) for each $u \in L$.
  \item Keep all edges $u \to v$ for $(u,v) \in E$ (capacity 1).
  \item Add sink $t$ with edge $v \to t$ (capacity 1) for each $v \in R$.
\end{itemize}

Run max-flow. Each unit of flow corresponds to one matched pair.
The maximum matching has size $|f^*|$.
\end{thmbox}

\subsection{Edge-disjoint paths via max flow}

\begin{thmbox}[title=Algorithm]
\textbf{Goal:} Maximum number of $s$-$t$ paths sharing no edge.

\textbf{Reduction:} Keep the original graph, set every edge capacity to~1.
Run max-flow from $s$ to $t$.

The value $|f^*|$ equals the maximum number of edge-disjoint paths.
By the max-flow min-cut theorem, it also equals the minimum number of edges whose
removal disconnects $s$ from $t$.
\end{thmbox}

\chapter{Quizzes}

\section{Quiz \#7:}
\subsection{Bipartite matching via max flow}
\begin{tcolorbox}[title=Bipartite matching via max flow]
\small
\ttfamily
\raggedright

Alors pour cet algorithme, c'est une application de l'algorithme max flow donc c'est le même code en soit mais le contexte et les règles du graphe vont changer. \\
Le but de ce problema esta de donc dde faire un bipartite matching.¿MAIS QUE ES UN BIPARTITE MATCHING? Alors, on va avoir deux set d'éléments et le but c'est de faire le maximum de paires possibles sachant que chaque élément ne peut faire qu'une paire qu'une fois avec un seul élément de l'autre team et que on sait que chaque élément ne peut se pairer que avec certains. \\

\vspace{2mm}
\textbf{Modélisation du graphe :}
\begin{center}
\begin{tabular}{c}
src $\rightarrow$ tous les éléments du premier set \\
éléments set 1 $\rightarrow$ éléments set 2 compatibles \\
éléments set 2 $\rightarrow$ end
\end{tabular}
\end{center}

\vspace{2mm}
et ensuite le max flow possible sur chaque branche est de 1 \\
Ainsi, chaque branche ne pourra être traversée que 1 fois et on ne va pas construire des doublons de paires ou des mauvaises paires.

\end{tcolorbox}

% =========================================================

\subsection{Edge disjoint paths via max flow}

\begin{tcolorbox}[title=Edge disjoint paths via max flow]
\small
\ttfamily
\raggedright

Alors pour ce second algorithme, c'est un peu différend mais c'est sur la même base, on va utiliser le max flow algorithm pour solver edge disjoint paths. ¿MAIS QUE ES UN EDGE DISJOINT PATHS? On cherche à trouver le maximum de chemins de A à B (de src to end) qui n'ont pas d'intersections entre eux donc qui passent par des passages uniques. \\

\vspace{2mm}
\textbf{Idée : modélisation en graphe}
\begin{center}
\begin{tabular}{c}
chaque chemin = arêtes du graphe \\
capacité de chaque arête = 1
\end{tabular}
\end{center}

\vspace{2mm}
Pour cela, on va modéliser en graphe notre réseau de chemins ect et par contre, comme précédemment, on va faire en sorte que le le max flow possible sur chaque branche est de 1. Ainsi, chaque branche ne pourra être traversée que 1 fois et on ne va pas l'utiliser en doublon et au final, le max flow sera le nombre de chemins que l'on a construits au maximum et qui sont disjoints.

\end{tcolorbox}

\end{document}
